---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

```{r , echo=TRUE, message =FALSE, warning=FALSE}
#install.packages('tidycensus')

library(tidycensus)
library(tidyverse)
library(data.table)
library(readxl)
#install.packages('maps')
library(maps)
library(stargazer)
#install.packages('R.utils')

library(R.utils)
options(tigris_use_cache = TRUE)
```


# Download census data


https://rconsortium.github.io/censusguide/r-packages-all.html

https://walker-data.com/tidycensus/articles/spatial-data.html


The following code uses the tidycensus package and the Census API to download data + shapefiles from the US Census.

- Declare Census product ( decennial).

- Get values for total population in 2010 (just for an example), by state. Get shapefiles by geometry=True

- Subset to US mainland; filter out Alaska, Puerto Rico, Hawaii. Convert state names to lowercase



`################
```{r}

#Declare parameters - API key, variables, time

census_api_key("2f1473c692f61175605ea04cbe2a9a1b41d5bf7c")

yr_select <- c(2010)

#call DF

stpop <- get_decennial(geography = "state", 
                        variables = "P003001", 
                         geometry = TRUE,
                        year = yr_select) %>% select(-variable) %>%
  filter(NAME!="Puerto Rico", NAME!="Alaska", NAME!="Hawaii" )
  
stpop$NAME <- tolower(stpop$NAME) 


###############################

```

#  Load and merge with analysis data

```{r , echo=TRUE, message=FALSE, warning=FALSE}

redist <- read.csv("./source_data/analysis_data.csv", stringsAsFactors = FALSE)


election_map<-merge(stpop, redist, by.x = "NAME", by.y = "state_full" )


```



## Maps - Redistricting policy

- Subset first to any year, say, 2012

-could make another map graphic that only highlights states that changed from 2000 to 2010



```{r, warning=FALSE}



#idvars = c("NAME", "GEOID", "geometry")

# Visualization

##Maps

map2012 <- election_map %>% filter(Year==2012)

## democratic votes

map2012 %>% 
  ggplot(aes(fill = Institution)) + 
  geom_sf(aes(color = Institution, fill = Institution))+
  labs(title = "Turnout percentage - 2012 Congressional Elections")+
  theme(plot.title =element_text(hjust=0.5, face="bold"), #moves title to the middle of graph, bolds it
          text= element_text(family= "Times", size = ))

```


## Descriptive stats

### Barplot/ boxplot of turnout

- once registered voters is merged, should look instead at percentage of eligible voters who  actually voted instead of just overall turnout


## Turnout - lineplot thru time, by institution

``` {r, echo=TRUE, message =FALSE, warning=FALSE}


### By state
election_map %>%
  ggplot(aes(x=Year, y = turnout_perc, colour = State)) + 
  geom_line(size = 0.5) + 
  labs(title = "Dem and Rep voter turnout from 2000-2020", x = "year", y = "percentage of vote by party") + 
  
  theme(plot.title = element_text(hjust = 0.5))


# Aggregate to institution
inst_lineplt <- election_map %>% group_by(Institution, Year) %>% summarise(avg_turnout = mean(turnout_perc))


inst_lineplt %>% 
   ggplot(aes(x=Year, y = avg_turnout, colour = Institution)) + 
  geom_line(size = .7) + 
  labs(title = "Avg. Turnout percentage 2000-2020, by redistricing ", x = "year", y = "Turnout percentage") +
    
  theme(plot.title = element_text(hjust = 0.5))



```

## Turnout - lineplot thru time, by institution - midterms vs presidential elections


``` {r, echo=TRUE, message =FALSE, warning=FALSE}

# Aggregate - midterm vs prez
midtm <- election_map %>% filter(mid_term==1) %>%
  group_by(Institution, Year) %>% 
  summarise(avg_turnout = mean(turnout_perc))

prez <- election_map %>% filter(mid_term==0) %>%
  group_by(Institution, Year) %>% 
  summarise(avg_turnout = mean(turnout_perc))


midtm %>% 
   ggplot(aes(x=Year, y = avg_turnout, colour = Institution)) + 
  geom_line(size = .7) + 
  labs(title = "Avg. Turnout - Midterms ", x = "year", y = "Turnout percentage") +
    
  theme(plot.title = element_text(hjust = 0.5))


prez %>% 
   ggplot(aes(x=Year, y = avg_turnout, colour = Institution)) + 
  geom_line(size = .7) + 
  labs(title = "Avg. Turnout -Pres. election, by redistricing ", x = "year", y = "Turnout percentage") +
    
  theme(plot.title = element_text(hjust = 0.5))


```



  
# Regression 

Our main independent variable derives from the 'Institution' or 'Drawn.By' columns. Several options:

1. Categorical/factor variable

2. Get dummies for each category (court, legislature, independent etc)

3. Make custom variables 'partisan'=1 or 0,  and categorize the above cases into it.


``` {r}
# not sure if it's appropriate to use state as a control
turnout_regression <- lm(turnout_perc ~ Institution+mid_term,
                         data = election_map)
summary(turnout_regression)
```

## Subset to states which actually changed policies


```{r}

redist_turn_short <- election_map %>%
  select(State, NAME, Year, Drawn.by, changed, Seats, turnout_perc,
         legislature, partisan, independent_commission, court, mid_term) %>%
  mutate(turnout_perc = turnout_perc / 100) %>%
  filter(Seats != 1,
         Year < 2022)

redist_turn_short %>%
  ggplot(aes(x = Drawn.by, y = turnout_perc)) +
  geom_point()

changed_subset <- redist_turn_short %>%
  filter(changed == 1) 

legislature_regression <- lm(turnout_perc ~ legislature + mid_term,
                         data = changed_subset)

court_regression <- lm(turnout_perc ~ court + mid_term,
                         data = changed_subset)

stargazer(legislature_regression, court_regression, type = "text")

```



## More maps/graphs

### Vote shares

This section adds the elections data, and performs basic cleaning/transformation to merge it with the previously made redistricting data frame. This produces a new data frame with district-level observations, and each row containing the number of votes for each candidate (Republican/Democrat/Other), as well as the redistricting method for that election.

Note: Minnesota's Democratic party is known as the "Democratic-Farmer-Labor" party and required special handling.


```{r}
elec <- read.csv("./source_data/1976-2020-house.csv")

votes <- elec %>%
  filter(year >= 2002) %>%
  mutate(party = ifelse(party == "REPUBLICAN", "REPUBLICAN",
                 ifelse(party == "DEMOCRAT", "DEMOCRAT",
                 ifelse(str_detect(party, "DEMOCRATIC-FARM"), "DEMOCRAT",
                        "OTHER")))) %>%
  group_by(year, state_po, district, party) %>%
  summarize(votes = sum(candidatevotes), totalvotes = mean(totalvotes)) %>%
  pivot_wider(names_from = party, values_from = votes, values_fill = 0) %>%
  left_join(redistricting, by = c("year"="Year", "state_po"="State"))

votes
```


The visualization filters the results to elections in states with more than one district, and also excludes races where a Democrat or Republican received zero votes (i.e. they were not on the ballot).

```{r}
votes %>%
  filter(DEMOCRAT > 0, REPUBLICAN > 0, Seats > 1) %>%
  mutate(winner = ifelse(REPUBLICAN > DEMOCRAT, "Republican", "Democrat")) %>%
  ggplot(aes(x = DEMOCRAT, y = REPUBLICAN)) +
  geom_point(alpha = 0.1) +
  scale_y_continuous(labels = label_comma(suffix = "k", scale = 0.001), limits = c(0, 300000)) +
  scale_x_continuous(labels = label_comma(suffix = "k", scale = 0.001), limits = c(0, 300000)) +
  facet_wrap(~Drawn.by)
```



### Proportion of states by institution




