---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

```{r , echo=TRUE, message =FALSE, warning=FALSE}
#install.packages('tidycensus')

library(tidycensus)
library(tidyverse)
library(data.table)
options(tigris_use_cache = TRUE)
```


#  load two datasets - election results, and redistricting

```{r , echo=TRUE, message =FALSE, warning=FALSE}
elec <- read.csv("/Users/Chuyuan/Downloads/1976-2020-house.csv", stringsAsFactors = FALSE)


redistricting_raw <- read.csv("/Users/Chuyuan/Downloads/StatesAndCyclesData.csv", stringsAsFactors = FALSE)
```

\

## Clean Redistricting data


Reading, dropping the state-level data, and also filling in blanks in the
Drawn.by column with values from the Institution column as a substitute.    
It is important to favor the Drawn.by column as it reflects court decisions.

we also need to identify the states that had their maps redrawn due to legal challenges. This resulted in them having more than 3 maps (rows) over the period

\

```{r , echo=TRUE, message =FALSE, warning=FALSE}
# Reading, dropping the state-level data, and also filling in blanks in the
# Drawn.by column with values from the Institution column as a substitute.
# It is important to favor the Drawn.by column as it reflects court decisions.
redistricting_raw <- redistricting_raw %>%
  filter(Level == "Congress") %>%
  mutate(Drawn.by = ifelse(Drawn.by == "", Institution, Drawn.by))


# Identifying the states that had their maps redrawn due to legal challenges
# This resulted in them having more than 3 maps (rows) over the period
problem_states <- redistricting_raw %>%
  count(State) %>%
  filter(n > 3) %>%
  select(State) %>%
  pull(State)
problem_states
```

\

Finally putting the puzzle together/trimming the excess map-election pairs:
 This...
1. Identifies problem states (probably not even necessary, but being thorough)
2. For problem states, finds the maps where:
 a. The start date for a map is after the election (not in effect yet), or
 b. the end date is before the election (superseded)

```{r , echo=TRUE, message =FALSE, warning=FALSE}
# Manually creating a DF to pair redistricting cycles with election years
# This will create a lot of extra cycle-year pairings for the aforementioned
# problem states--pairing them with elections before the maps were drawn and
# also after the maps were sent back to the legislature or redrawn by a court.
cycle_pairs <- rbind(
  data.frame(Cycle.Year = 2000, Year = c(2002, 2004, 2006, 2008, 2010)),
  data.frame(Cycle.Year = 2010, Year = c(2012, 2014, 2016, 2018, 2020)),
  data.frame(Cycle.Year = 2020, Year = 2022))

# Joining the dataframes, creating a dummy variable for year comparisons,
# and filling in blank values for the dates with an arbitrary date (missing
# data was causing errors for the difftime function)
redistricting_joined <- full_join(redistricting_raw, cycle_pairs, by = "Cycle.Year") %>%
  mutate(Election.Cutoff = paste(Year, "-11-01", sep = ""),
         Start.Date = ifelse((Start.Date == "") & (Seats == 1), "2001-12-31", Start.Date),
         End.Date = ifelse((End.Date == "") & (Seats == 1), "2001-12-31", End.Date))

# Finally putting the puzzle together/trimming the excess map-election pairs:
# This...
# 1. Identifies problem states (probably not even necessary, but being thorough)
# 2. For problem states, finds the maps where:
#     a. The start date for a map is after the election (not in effect yet), or
#     b. the end date is before the election (superseded), then...
# 3. Marks those maps as inactive == TRUE, to then be filtered out.
redistricting <- redistricting_joined %>%
  mutate(Inactive = ifelse(!(State %in% problem_states), F,
                    ifelse(difftime(redistricting_joined$Start.Date, Election.Cutoff) > 0 |
                           difftime(redistricting_joined$End.Date, Election.Cutoff) < 0, T, F))) %>%
  filter(Inactive == FALSE)

# Example of what all this nonsense was for
redistricting %>%
  filter(State %in% problem_states) %>%
  select(State, Drawn.by, Year, Start.Date, End.Date)
```

```{r , echo=TRUE, message = FALSE, warning=FALSE}
# Adding various dichotomous dummy variables (1 for mathces, 0 otherwise)
# legislature: Drawn.by == "Legislature"
# partisan: Drawn by legislature or politician commission
# independent_commission: Drawn.by == "Independent commission"
# court: Drawn.by contains "court"
# midterm: Not a presidential election year
#
# to possibly add later:
#   Whether a state changed to/from a non-partisan method
#   Whether or not the state had a US Senate election that year

redistricting <- redistricting %>%
  mutate(legislature = ifelse(Drawn.by == "Legislature", 1, 0),
         partisan = ifelse((Drawn.by == "Legislature") |
                           (Drawn.by == "Politician commission") , 1, 0),
         independent_commission = ifelse(Drawn.by == "Independent commission", 1, 0),
         court = ifelse(str_detect(tolower(Drawn.by), "court"), 1, 0),
         mid_term = ifelse(Year %% 4 == 2, 1, 0))

redistricting %>%
  select(Year, State, Drawn.by, legislature, partisan, independent_commission, court, mid_term)


```

\

## Clean Election data 


Keep only for required span - 2000 to 2020

Drop states with only one Congressional district. 

Then, convert the party variable to have only three categories - Republican, Democrat and Other. Pivot to wide, and get vote totals and shares.

Groupby/ aggregate and summarize district-level data to state level data.

\

```{r , echo=TRUE, message =FALSE, warning=FALSE}
# Keep only for required span - 2000 to 2020
  # drop  single dist states  
    
  # Vary from year?
single_dists <- c("AK", "DE", "MT", "ND", "SD", "VT", "WY")

elec <- elec %>%
  filter(year >= 2000) %>%
  filter(!state_po %in% single_dists)





############################################################
# Aggregation

# Gen new party2 variable

elec <- elec %>%
  mutate(party2 = case_when(
    party =="DEMOCRAT" ~ 'DEMOCRAT',
    party=="REPUBLICAN" ~ "REPUBLICAN",
    TRUE ~ as.character('OTHER'))) 

# need GEOID to merge with census map
#elec$GEOID <- paste(elec$state_fips, elec$district, sep = "")

# Pivot wide - column for party totals
votes <- elec %>% pivot_wider(names_from = party2, values_from = candidatevotes) %>%
  mutate(across(c(DEMOCRAT, REPUBLICAN, OTHER), ~case_when(.== "NULL" ~ 0, 
                                  TRUE ~as.numeric(as.character(.)))))

# total turnout column
votes$turnout <- votes$DEMOCRAT+votes$REPUBLICAN+votes$OTHER

# Data now is at State-Year-Many districts - 
# aggregate statistics for all districts to one observation for state

votes <- votes %>%  group_by(state_po, state, year) %>%
  summarize(demv = sum(DEMOCRAT), repv = sum(REPUBLICAN), othv = sum(OTHER), turnout = sum(turnout))

# create new "map_year" column in elec df as year-1 rounded to 10s
# the year in which the map for that election was drawn
votes <- votes %>%
  mutate(map_year = case_when(
    year <= 2010 ~ 2000,
   year>2010 ~ 2010))
```

# Merge 


Merge the *Redistricting* and *Votes* datasets, on two keys: State and Map-Year

```{r , echo=TRUE, message =FALSE, warning=FALSE}

df <- votes %>%   left_join(redistricting, by = c("state_po" = "State", "map_year" = "Cycle.Year"))


# Just cleaning it a bit for regression


df<- df %>% mutate(othv2=ifelse(is.na(othv), 0, othv))
                  
df$turnout2 <- df$othv2 + df$demv + df$repv
#df
```


# Download census data


https://rconsortium.github.io/censusguide/r-packages-all.html

https://walker-data.com/tidycensus/articles/spatial-data.html


The following code uses the tidycensus package and the Census API to download data + shapefiles from the US Census.

- Declare Census product (ACS, decennial), variables, time and geography.

- Reshape to wide, rename variables.

- Subset to US mainland; filter out Alaska, Puerto Rico, Hawaii.

- Generate a variable of percent black

- Plot map


```{r , echo=TRUE, message =FALSE, warning=FALSE}

# Declare parameters - API key, variables, time

census_api_key("YOUR KEY")

race_vars <- c("P003001", "P003002", "P003003")

yr_select <- c(2010)

# call DF

allst <- get_decennial(geography = "state", 
                        variables = race_vars, 
                         geometry = TRUE,
                        year = yr_select) 



#idvars = c("NAME", "GEOID", "geometry")

# SPREAD to wide
allst2 <- spread(allst, variable, value)

setnames(allst2, old = c("P003001", "P003002", "P003003"), new = c('tot_pop','white_pop','black_pop'))

allst2 <- allst2 %>% filter(NAME!="Puerto Rico", NAME!="Alaska", NAME!="Hawaii" )

allst2$black_perc <- allst2$black_pop/allst2$tot_pop


allst2 %>% ggplot(aes(fill = black_perc)) + geom_sf(color = NA) + 
  scale_fill_viridis_c(option = "viridis") 
```


## Join Eligible voters data 

# Descriptive results


Redistricting MAP

Vote shares and competitiveness - MAP/LINEPLOT

Census characteristics - AVERAGES?

```{r}
## democratic votes
election_map2<-merge(allst2, df, by.x = "NAME", by.y = "state" )
election_map2 %>% mutate(dpct = demv/turnout) %>%
  ggplot() + 
  geom_sf(aes(color = dpct, fill = dpct)) 
```

```{r , echo=TRUE, message =FALSE, warning=FALSE}
## Republican votes
election_map2 %>% mutate(rpct = repv/turnout) %>%
  ggplot(aes(fill = rpct)) + 
  geom_sf() +
  scale_fill_viridis_c(option = "viridis")
```

```{r, echo=TRUE, message =FALSE, warning=FALSE}

## This only shows 4 instutions
#election_map2<-merge(allst2, df, by.x = "NAME", by.y = "state" )
#election_map2 %>% group_by(map_year = "2000") %>%
  #ggplot(aes(fill = Institution)) + geom_sf(color = NA)  

## 

#install.packages("mapproj")
library(maps)
library(R.utils)
usa <- map_data("state")
df$state<- tolower(df$state)
df$state<- capitalize(df$state)
election_map2 <- merge(usa, df, by.x = "region", by.y = "state")

## 2000-2010
election_map2 %>%
  group_by(map_year = "2000") %>%
  arrange(group, order) %>%
  ggplot(aes(x = long, y = lat, group = group, fill = Institution)) +
  geom_polygon(colour = "black") +
  coord_map("polyconic")

## 2010-2020
election_map2 %>%
  group_by(map_year = "2010") %>%
  arrange(group, order) %>%
  ggplot(aes(x = long, y = lat, group = group, fill = Institution)) +
  geom_polygon(colour = "black") +
  coord_map("polyconic")
```


# Regressions

Turnout percent into 

``` {r, echo=TRUE, message =FALSE, warning=FALSE}
## create turnout_df
x <- aggregate(turnout ~ year, votes, sum)
y <- aggregate(demv ~ year, votes, sum)
z <-aggregate(repv ~ year, votes, sum)
turnout_df <- data.frame(x,y,z)
turnout_df <- select(turnout_df, -year.1, -year.2)

## combine demv and repv in one column
df2 <- pivot_longer(turnout_df, demv:repv, names_to = "party", values_to = "partyvotes")

## create turnout percent visual
df2 %>%
  mutate(vote_pct = partyvotes/turnout) %>%
  ggplot(aes(x=year, y = vote_pct, colour = party)) + 
  geom_line(size = 0.5) + 
  labs(title = "Dem and Rep voter turnout from 2000-2020", x = "year", y = "percentage of vote by party") + 
  scale_colour_manual(values = c("#56B4E9", "#FF9999"),
                    guide = guide_legend(reverse = TRUE),
                    labels=c("dem_pct","rep_pct")) +
  theme(plot.title = element_text(hjust = 0.5)) 
```


```{r, echo=TRUE, message =FALSE, warning=FALSE}
## create pie chart show Institution proportion in 2000-2010
library(scales)
count <- df %>% group_by(map_year) %>%
       filter(map_year == "2000") %>%
       count(Institution)
count %>% 
  mutate(prop = n/sum(n)) %>%
  ggplot(aes(x="", y=n, fill=Institution)) +
  geom_bar(stat="identity", width=1, color="black") +
  coord_polar("y", start=0) +
  geom_text(aes(x=1.4, label=percent(prop)), position=position_stack(vjust=0.5), size=3) +
  labs(x = NULL, y = NULL) +
  scale_fill_brewer(palette="Pastel2")

## create pie chart show Institution proportion in 2010-2020
count <- df %>% group_by(map_year) %>%
       filter(map_year == "2010") %>%
       count(Institution)
count %>% 
  mutate(prop = n/sum(n)) %>%
  ggplot(aes(x="", y=n, fill=Institution)) +
  geom_bar(stat="identity", width=1, color="black") +
  coord_polar("y", start=0) +
  geom_text(aes(x=1.4, label=percent(prop)), position=position_stack(vjust=0.5), size=2) +
  labs(x = NULL, y = NULL) +
  scale_fill_brewer(palette="Pastel2")
```

```{r,echo=TRUE, message =FALSE, warning=FALSE}
## The effect of each institution on the share of the Democratic vote
df %>%
  mutate(dpct = demv/turnout) %>%
  ggplot(aes(x= year, y = dpct, colour = Drawn.by)) +
      geom_point() +
      labs(title="Title")
## Major contributors to encouraging Democratic votes are legislative, politician committees, state courts (2000-2010),andindependent commision (2010-2020).


## The effect of each institution on the share of the Republican vote
df %>%
  mutate(rpct = repv/turnout) %>%
  ggplot(aes(x= year, y = rpct, colour = Drawn.by)) +
      geom_point() +
      labs(title="Title")
## Major contributors to encouraging Republican votes are legislative, Federal court, and state courts,and independent commision.
```
