---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

```{r , echo=TRUE, messages=FALSE, warning=FALSE}
#install.packages('tidycensus')

library(tidycensus)
library(tidyverse)
library(data.table)
library(readxl)
library(stargazer)
library(scales)
library(dplyr)

options(tigris_use_cache = TRUE)

```


# redistricting and census voting




\

## Clean Redistricting data


Reading, dropping the state-level data, and also filling in blanks in the
Drawn.by column with values from the Institution column as a substitute.    
It is important to favor the Drawn.by column as it reflects court decisions.

we also need to identify the states that had their maps redrawn due to legal challenges. This resulted in them having more than 3 maps (rows) over the period

\
```{r , echo=TRUE, message =FALSE, warning=FALSE}


#elec <- read.csv("./source_data/1976-2020-house.csv", stringsAsFactors = FALSE)


```

# Data Cleaning and Merging

We worked with three separate data sets for our research project, which needed to be organized and merged to state-election year pairings for our analysis. Single district states were left in the data to be included for certain visualizations, but they would ultimately be dropped for most of the substantive analysis.

## Redistricting Data

This data set began as state-redistricting map pairings, tied to the original census year that maps were based on. A significant complicating factor, however, is that several states have map revisions as a result of court challenges, creating inconsistent numbers of pairings. Fortunately, the data also has columns denoting the years a given map was active, which could be used to associate an individual map with the elections where it was active. Since this data frame would be merged with both of our other data sets, this is also also the step where we added dichotomous dummy variables for use in later analysis.

This was merged in several steps:
1. Filtering out state house and state senate maps, as well as the 2020 congressional maps
2. Filling in blank entries in the "Drawn.by" column with "Institution"
3. Identifying states that have revised maps (specifically FL, NC, OH, PA, RI, TX, VA)
4. Creating an additional data frame for pairing census cycles to election years
5. Doing a full_join on the data frames to create all possible pairings
6. Adding a dummy variable for November 1st of each year as a cutoff point to mark election dates
7. Using the above to create another dummy variable to mark election year-map pairings where a map was inactive and filter them out
8. Adding a dummy variable for which states used more than one redistricting method in the sample window
9. Adding a variety of dichotomous dummy variables to identify different redistricting methods


```{r , echo=TRUE, message =FALSE, warning=FALSE}
redistricting_raw <- read.csv("./source_data/StatesAndCyclesData.csv", stringsAsFactors = FALSE)

# 1. Filtering out state house and state senate maps, as well as the 2020 congressional maps
# 2. Filling in blank entries in the "Drawn.by" column with "Institution"
redistricting_raw <- redistricting_raw %>%
  filter(Level == "Congress", Cycle.Year < 2020) %>%
  mutate(Drawn.by = ifelse(Drawn.by == "", Institution, Drawn.by))

# 3. Identifying states that have revised maps (FL, NC, OH, PA, RI, TX, VA)
problem_states <- redistricting_raw %>%
  count(State) %>%
  filter(n > 2) %>%
  select(State) %>%
  pull(State)

# 4. Creating an additional data frame for pairing census cycles to election years
cycle_pairs <- rbind(
  data.frame(Cycle.Year = 2000, Year = c(2002, 2004, 2006, 2008, 2010)),
  data.frame(Cycle.Year = 2010, Year = c(2012, 2014, 2016, 2018, 2020)))

# 5. Doing a full_join on the data frames to create all possible pairings
# 6. Adding a dummy variable for November 1st of each year as a cutoff point to mark election dates
redistricting_joined <- full_join(redistricting_raw, cycle_pairs, by = "Cycle.Year") %>%
  mutate(Election.Cutoff = paste(Year, "-11-01", sep = ""),
         Start.Date = ifelse((Start.Date == "") & (Seats == 1), "2001-12-31", Start.Date),
         End.Date = ifelse((End.Date == "") & (Seats == 1), "2001-12-31", End.Date))

# 7. Using the above to create another dummy variable to mark election year-map pairings where a map was inactive and filter them out
redistricting <- redistricting_joined %>%
  mutate(Inactive = ifelse(!(State %in% problem_states), F,
                    ifelse(difftime(redistricting_joined$Start.Date, Election.Cutoff) > 0 |
                           difftime(redistricting_joined$End.Date, Election.Cutoff) < 0, T, F))) %>%
  filter(Inactive == FALSE)

# 8. Adding a dummy variable for which states used more than one redistricting method in the sample window
redistricting_changes = redistricting %>%
  group_by(State) %>%
  summarize(redistricting_methods = n_distinct(Drawn.by)) %>%
  mutate(changed = ifelse(redistricting_methods > 1, 1, 0))

redistricting <- left_join(redistricting, redistricting_changes)

# Example showing the successful isolation of map revisions and accurate pairings
redistricting %>%
  filter(State %in% problem_states) %>%
  select(State, Drawn.by, Year, Start.Date, End.Date, changed)

# 9. Adding a variety of dichotomous dummy variables to identify different redistricting methods for later analysis
redistricting <- redistricting %>%
  mutate(legislature = ifelse(Drawn.by == "Legislature", 1, 0),
         partisan = ifelse((Drawn.by == "Legislature") |
                           (Drawn.by == "Politician commission") , 1, 0),
         independent_commission = ifelse(Drawn.by == "Independent commission", 1, 0),
         court = ifelse(str_detect(tolower(Drawn.by), "court"), 1, 0),
         mid_term = ifelse(Year %% 4 == 2, 1, 0))

redistricting %>%
  select(Year, State, Drawn.by, changed, legislature, partisan,
         independent_commission, court, mid_term)

```

## Election results

The MIT election lab data was more granular than necessary for our analysis, containing state-election year-district-candidate observations with vote totals for each candidate. It was aggregated to the district level, pivoted wider to create state-election year-district observations, and then merged with the redistricting data. Third party candidates were grouped as "OTHER" for the aggregation, and Minnesota's Democratic party was renamed to match the national party.

```{r}
elec <- read.csv("./source_data/1976-2020-house.csv")

votes <- elec %>%
  filter(year >= 2002) %>%
  mutate(party = ifelse(party == "REPUBLICAN", "REPUBLICAN",
                 ifelse(party == "DEMOCRAT", "DEMOCRAT",
                 ifelse(str_detect(party, "DEMOCRATIC-FARM"), "DEMOCRAT",
                        "OTHER")))) %>%
  group_by(year, state_po, district, party) %>%
  summarize(votes = sum(candidatevotes), totalvotes = mean(totalvotes)) %>%
  pivot_wider(names_from = party, values_from = votes, values_fill = 0) %>%
  left_join(redistricting, by = c("year"="Year", "state_po"="State")) %>% 
  mutate(dem_margin = 100*(DEMOCRAT-REPUBLICAN)/totalvotes)

write.csv(votes, "votes_redistricting.csv", row.names=FALSE)
votes
```

\

## Census Data 
This data set contained State-election year-voter 
Keep only for required span - 2000 to 2020
Drop states with only one Congressional district. 
Then, convert the party variable to have only three categories - Republican, Democrat and Other. Pivot to wide, and get vote totals and shares.


Census data was available for the current census and historical data (separated).
Current data is combined, historical data is by presidential/midterms.
We import the relevant cells from the Excel sheets and merge them as needed.


```{r, echo=FALSE, messages=FALSE}
states <- elec %>%
    mutate("state_full" = tolower(state)) %>%
    select("state_full", "state_po") %>%
    distinct(state_full, state_po)

# 2020 file has raw populations (others have only percentages)
raw_2020 <- read_excel("./source_data/VoteAndRegByState2020.xlsx", range = "A7:E58")

names(raw_2020) <- c("State", "population", "citizens", "registered", "registered_percent")

# Presidential years (in 'broken' format in excel sheet (printer friendly))
rng <- "A7:K58"
prez_wide_regA <- read_excel("./source_data/VoteAndRegByStateHistory_prez_reg.xlsx", range=rng)
rng <- "B74:C125"
prez_wide_regB <- read_excel("./source_data/VoteAndRegByStateHistory_prez_reg.xlsx", range=rng)

prez_wide <- data.frame(cbind(prez_wide_regA, prez_wide_regB)[,seq(1,14,2)])
names(prez_wide) <- c("turnout_state", c(2020, 2016, 2012, 2008, 2004, 2000))

# MidTerms years (not in 'broken' format)
rng <- "A7:K58"
midt_wide_reg <- read_excel("./source_data/VoteAndRegByStateHistory_congress.xlsx", range=rng)

midt_wide <- data.frame(midt_wide_reg[,seq(1,12,2)])
names(midt_wide) <- c("turnout_state", c(2018, 2014, 2010, 2006, 2002))

all_years <- left_join(prez_wide, midt_wide,
                       by = c("turnout_state" = "turnout_state"))

all_years_long <- pivot_longer(all_years, cols=colnames(all_years)[-1], names_to="turnout_year", values_to="turnout_perc")

all_years_long <- all_years_long %>%
  mutate(turnout_year = as.numeric(turnout_year),
         turnout_state = tolower(turnout_state))

```


# MERGE
```{r}
redistricting <- left_join(redistricting, states,
                           by = c("State" = "state_po")) %>%
                 mutate(Cycle.Year = as.numeric(Cycle.Year),
                        state_full = tolower(state_full))

redist_turn <- left_join(redistricting, all_years_long,
                         by = c("state_full" = "turnout_state", "Year" = "turnout_year"))
```


```{r}
redist_turn_short <- redist_turn %>%
  select(State, state_full, Year, Drawn.by, changed, Seats, turnout_perc,
         legislature, partisan, independent_commission, court, mid_term) %>%
  mutate(turnout_perc = turnout_perc / 100) %>%
  filter(Seats != 1,
         Year < 2022)

redist_turn_short %>%
  ggplot(aes(x = Drawn.by, y = turnout_perc)) +
  geom_point()

```
```{r}
# not sure if it's appropriate to use state as a control
turnout_regression <- lm(turnout_perc ~ legislature + State + mid_term,
                         data = redist_turn_short[redist_turn_short$Seats > 1,])
summary(turnout_regression)
```


```{r}
changed_subset <- redist_turn_short %>%
  filter(changed == 1) 

legislature_regression <- lm(turnout_perc ~ legislature + mid_term,
                         data = changed_subset)

court_regression <- lm(turnout_perc ~ court + mid_term,
                         data = changed_subset)

stargazer(legislature_regression, court_regression, type = "text")

```


```{r}
court_states <- redist_turn_short %>%
  filter(court == 1) %>%
  distinct(State) %>%
  pull()

legislature_states <- redist_turn_short %>%
  filter(legislature == 1) %>%
  distinct(State) %>%
  pull()

court_subset <- redist_turn_short %>%
  filter(changed == 1, State %in% court_states, State %in% legislature_states)

court_regression2 <- lm(turnout_perc ~ court + mid_term,
                        data = changed_subset)

summary(court_regression2)
```


The visualization filters the results to elections in states with more than one district, and also excludes races where a Democrat or Republican received zero votes (i.e. they were not on the ballot).

```{r}
votes %>%
  filter(DEMOCRAT > 0, REPUBLICAN > 0, Seats > 1) %>%
  mutate(winner = ifelse(REPUBLICAN > DEMOCRAT, "Republican", "Democrat")) %>%
  ggplot(aes(x = DEMOCRAT, y = REPUBLICAN)) +
  geom_point(alpha = 0.1) +
  geom_line(data = data.frame(DEMOCRAT = c(0, 300000), REPUBLICAN = c(0, 300000))) +
  scale_y_continuous(labels = label_comma(suffix = "k", scale = 0.001), limits = c(0, 300000)) +
  scale_x_continuous(labels = label_comma(suffix = "k", scale = 0.001), limits = c(0, 300000)) +
  facet_wrap(~Drawn.by)
# write comment: "Diagonal represents tipping point of victory towards party"
```
```{r}



# Create a new data frame with the avg_margin values grouped by year and Drawn.by

inst_lineplt <- votes %>% group_by(Drawn.by, year) %>% summarise(avg_margin = mean(dem_margin))


inst_lineplt %>% 
   ggplot(aes(x=year, y = avg_margin, colour = Drawn.by)) + 
  geom_line(size = .7) + 
  labs(title = "Vote share margins, by redistricting ", x = "year", y = "Dem Margin in (%)") +
    
  theme(plot.title = element_text(hjust = 0.5))

```
\




```{r}
votes %>% filter(Drawn.by=="Legislature") %>%
  filter(DEMOCRAT > 0, REPUBLICAN > 0, Seats > 1) %>%
  ggplot(aes(x = dem_margin)) +
    geom_histogram(binwidth = 3) +
    geom_vline(xintercept = 0, color = "black") +
    labs(title = "Democratic vote margin (Legislature)",
         x = "Margin (percent)",
         y = "Frequency")
```
```

