---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

```{r , echo=TRUE, messages=FALSE, warning=FALSE}
#install.packages('tidycensus')

library(tidycensus)
library(tidyverse)
library(data.table)
library(readxl)
options(tigris_use_cache = TRUE)

```


# redistricting and census voting

```{r , echo=TRUE, message =FALSE, warning=FALSE}

redistricting_raw <- read.csv("./source_data/StatesAndCyclesData.csv", stringsAsFactors = FALSE)

elec <- read.csv("./source_data/1976-2020-house.csv", stringsAsFactors = FALSE)
states <- elec %>%
    mutate("state_full" = tolower(state)) %>%
    select("state_full", "state_po") %>%
    distinct(state_full, state_po)

```

\

## Clean Redistricting data


Reading, dropping the state-level data, and also filling in blanks in the
Drawn.by column with values from the Institution column as a substitute.    
It is important to favor the Drawn.by column as it reflects court decisions.

we also need to identify the states that had their maps redrawn due to legal challenges. This resulted in them having more than 3 maps (rows) over the period

\

```{r , echo=TRUE, message =FALSE, warning=FALSE}
# Reading, dropping the state-level data, and also filling in blanks in the
# Drawn.by column with values from the Institution column as a substitute.
# It is important to favor the Drawn.by column as it reflects court decisions.
redistricting_raw <- redistricting_raw %>%
  filter(Level == "Congress") %>%
  mutate(Drawn.by = ifelse(Drawn.by == "", Institution, Drawn.by))


# Identifying the states that had their maps redrawn due to legal challenges
# This resulted in them having more than 3 maps (rows) over the period
problem_states <- redistricting_raw %>%
  count(State) %>%
  filter(n > 3) %>%
  select(State) %>%
  pull(State)
problem_states
```

\

Finally putting the puzzle together/trimming the excess map-election pairs:
 This...
1. Identifies problem states (probably not even necessary, but being thorough)
2. For problem states, finds the maps where:
 a. The start date for a map is after the election (not in effect yet), or
 b. the end date is before the election (superseded)

```{r , echo=TRUE, message =FALSE, warning=FALSE}
# Manually creating a DF to pair redistricting cycles with election years
# This will create a lot of extra cycle-year pairings for the aforementioned
# problem states--pairing them with elections before the maps were drawn and
# also after the maps were sent back to the legislature or redrawn by a court.
cycle_pairs <- rbind(
  data.frame(Cycle.Year = 2000, Year = c(2002, 2004, 2006, 2008, 2010)),
  data.frame(Cycle.Year = 2010, Year = c(2012, 2014, 2016, 2018, 2020)),
  data.frame(Cycle.Year = 2020, Year = 2022))

# Joining the dataframes, creating a dummy variable for year comparisons,
# and filling in blank values for the dates with an arbitrary date (missing
# data was causing errors for the difftime function)
redistricting_joined <- full_join(redistricting_raw, cycle_pairs, by = "Cycle.Year") %>%
  mutate(Election.Cutoff = paste(Year, "-11-01", sep = ""),
         Start.Date = ifelse((Start.Date == "") & (Seats == 1), "2001-12-31", Start.Date),
         End.Date = ifelse((End.Date == "") & (Seats == 1), "2001-12-31", End.Date))

#view(redistricting_joined)

# Finally putting the puzzle together/trimming the excess map-election pairs:
# This...
# 1. Identifies problem states (probably not even necessary, but being thorough)
# 2. For problem states, finds the maps where:
#     a. The start date for a map is after the election (not in effect yet), or
#     b. the end date is before the election (superseded), then...
# 3. Marks those maps as inactive == TRUE, to then be filtered out.
redistricting <- redistricting_joined %>%
  mutate(Inactive = ifelse(!(State %in% problem_states), F,
                    ifelse(difftime(redistricting_joined$Start.Date, Election.Cutoff) > 0 |
                           difftime(redistricting_joined$End.Date, Election.Cutoff) < 0, T, F))) %>%
  filter(Inactive == FALSE)

# Example of what all this nonsense was for
redistricting %>%
  filter(State %in% problem_states) %>%
  select(State, Drawn.by, Year, Start.Date, End.Date)

```

```{r , echo=TRUE, message = FALSE, warning=FALSE}
# Adding various dichotomous dummy variables (1 for mathces, 0 otherwise)
# legislature: Drawn.by == "Legislature"
# partisan: Drawn by legislature or politician commission
# independent_commission: Drawn.by == "Independent commission"
# court: Drawn.by contains "court"
# midterm: Not a presidential election year
#
# to possibly add later:
#   Whether a state changed to/from a non-partisan method
#   Whether or not the state had a US Senate election that year

redistricting <- redistricting %>%
  mutate(legislature = ifelse(Drawn.by == "Legislature", 1, 0),
         partisan = ifelse((Drawn.by == "Legislature") |
                           (Drawn.by == "Politician commission") , 1, 0),
         independent_commission = ifelse(Drawn.by == "Independent commission", 1, 0),
         court = ifelse(str_detect(tolower(Drawn.by), "court"), 1, 0),
         mid_term = ifelse(Year %% 4 == 2, 1, 0))

redistricting %>%
  select(Year, State, Drawn.by, legislature, partisan, independent_commission, court, mid_term)

# view(redistricting)
```

\

## Clean Election data 


Keep only for required span - 2000 to 2020

Drop states with only one Congressional district. 

Then, convert the party variable to have only three categories - Republican, Democrat and Other. Pivot to wide, and get vote totals and shares.


Census data was available for the current census and historical data (separated).
Current data is combined, historical data is by presidential/midterms.
We import the relevant cells from the Excel sheets and merge them as needed.


```{r, echo=FALSE, messages=FALSE}

# 2020 file has raw populations (others have only percentages)
raw_2020 <- read_excel("./source_data/VoteAndRegByState2020.xlsx", range = "A7:E58")

names(raw_2020) <- c("State", "population", "citizens", "registered", "registered_percent")

# Presidential years (in 'broken' format in excel sheet (printer friendly))
rng <- "A7:K58"
prez_wide_regA <- read_excel("./source_data/VoteAndRegByStateHistory_prez_reg.xlsx", range=rng)
rng <- "B74:C125"
prez_wide_regB <- read_excel("./source_data/VoteAndRegByStateHistory_prez_reg.xlsx", range=rng)

prez_wide <- data.frame(cbind(prez_wide_regA, prez_wide_regB)[,seq(1,14,2)])
names(prez_wide) <- c("turnout_state", c(2020, 2016, 2012, 2008, 2004, 2000))

# MidTerms years (not in 'broken' format)
rng <- "A7:K58"
midt_wide_reg <- read_excel("./source_data/VoteAndRegByStateHistory_congress.xlsx", range=rng)

midt_wide <- data.frame(midt_wide_reg[,seq(1,12,2)])
names(midt_wide) <- c("turnout_state", c(2018, 2014, 2010, 2006, 2002))

all_years <- left_join(prez_wide, midt_wide,
                       by = c("turnout_state" = "turnout_state"))

all_years_long <- pivot_longer(all_years, cols=colnames(all_years)[-1], names_to="turnout_year", values_to="turnout_perc")

all_years_long <- all_years_long %>%
  mutate(turnout_year = as.numeric(turnout_year),
         turnout_state = tolower(turnout_state))

```


# MERGE
```{r}

redistricting <- left_join(redistricting, states,
                           by = c("State" = "state_po")) %>%
                 mutate(Cycle.Year = as.numeric(Cycle.Year),
                        state_full = tolower(state_full))

redist_turn <- left_join(redistricting, all_years_long,
                         by = c("state_full" = "turnout_state", "Cycle.Year" = "turnout_year"))

```
