---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

```{r , echo=TRUE, message =FALSE, warning=FALSE}
#install.packages('tidycensus')

library(tidycensus)
library(tidyverse)
library(data.table)
library(readxl)
options(tigris_use_cache = TRUE)
```


#  load two datasets - election results, and redistricting

```{r , echo=TRUE, message =FALSE, warning=FALSE}
elec <- read.csv("../1976-2020-house.csv", stringsAsFactors = FALSE)

redistricting_raw <- read.csv("../StatesAndCyclesData.csv", stringsAsFactors = FALSE)
```

\

## Clean Redistricting data


Reading, dropping the state-level data, and also filling in blanks in the
Drawn.by column with values from the Institution column as a substitute.    
It is important to favor the Drawn.by column as it reflects court decisions.

we also need to identify the states that had their maps redrawn due to legal challenges. This resulted in them having more than 3 maps (rows) over the period

\

```{r , echo=TRUE, message =FALSE, warning=FALSE}
# Reading, dropping the state-level data, and also filling in blanks in the
# Drawn.by column with values from the Institution column as a substitute.
# It is important to favor the Drawn.by column as it reflects court decisions.
redistricting_raw <- redistricting_raw %>%
  filter(Level == "Congress") %>%
  mutate(Drawn.by = ifelse(Drawn.by == "", Institution, Drawn.by))


# Identifying the states that had their maps redrawn due to legal challenges
# This resulted in them having more than 3 maps (rows) over the period
problem_states <- redistricting_raw %>%
  count(State) %>%
  filter(n > 3) %>%
  select(State) %>%
  pull(State)

problem_states

```

\

Finally putting the puzzle together and trimming the excess map-election pairs:
 This...
1. Identifies problem states (probably not even necessary, but being thorough)
2. For problem states, finds the maps where:
 a. The start date for a map is after the election (not in effect yet), or
 b. the end date is before the election (superseded)

```{r , echo=TRUE, message =FALSE, warning=FALSE}
# Manually creating a DF to pair redistricting cycles with election years
# This will create a lot of extra cycle-year pairings for the aforementioned
# problem states--pairing them with elections before the maps were drawn and
# also after the maps were sent back to the legislature or redrawn by a court.
cycle_pairs <- rbind(
  data.frame(Cycle.Year = 2000, Year = c(2002, 2004, 2006, 2008, 2010)),
  data.frame(Cycle.Year = 2010, Year = c(2012, 2014, 2016, 2018, 2020)),
  data.frame(Cycle.Year = 2020, Year = 2022))

# Joining the dataframes, creating a dummy variable for year comparisons,
# and filling in blank values for the dates with an arbitrary date (missing
# data was causing errors for the difftime function)
redistricting_joined <- full_join(redistricting_raw, cycle_pairs, by = "Cycle.Year") %>%
  mutate(Election.Cutoff = paste(Year, "-11-01", sep = ""),
         Start.Date = ifelse((Start.Date == "") & (Seats == 1), "2001-12-31", Start.Date),
         End.Date = ifelse((End.Date == "") & (Seats == 1), "2001-12-31", End.Date))

# Finally putting the puzzle together/trimming the excess map-election pairs:
# This...
# 1. Identifies problem states (probably not even necessary, but being thorough)
# 2. For problem states, finds the maps where:
#     a. The start date for a map is after the election (not in effect yet), or
#     b. the end date is before the election (superseded), then...
# 3. Marks those maps as inactive == TRUE, to then be filtered out.
redistricting <- redistricting_joined %>%
  mutate(Inactive = ifelse(!(State %in% problem_states), F,
                    ifelse(difftime(redistricting_joined$Start.Date, Election.Cutoff) > 0 |
                           difftime(redistricting_joined$End.Date, Election.Cutoff) < 0, T, F))) %>%
  filter(Inactive == FALSE)

# Example of what all this nonsense was for
redistricting %>%
  filter(State %in% problem_states) %>%
  select(State, Drawn.by, Year, Start.Date, End.Date)
```

```{r , echo=TRUE, message =FALSE, warning=FALSE}
# Adding various dichotomous dummy variables (1 for mathces, 0 otherwise)
# legislature: Drawn.by == "Legislature"
# partisan: Drawn by legislature or politician commission
# independent_commission: Drawn.by == "Independent commission"
# court: Drawn.by contains "court"
# midterm: Not a presidential election year
#
# to possibly add later:
#   Whether a state changed to/from a non-partisan method
#   Whether or not the state had a US Senate election that year

redistricting <- redistricting %>%
  mutate(legislature = ifelse(Drawn.by == "Legislature", 1, 0),
         partisan = ifelse((Drawn.by == "Legislature") |
                           (Drawn.by == "Politician commission") , 1, 0),
         independent_commission = ifelse(Drawn.by == "Independent commission", 1, 0),
         court = ifelse(str_detect(tolower(Drawn.by), "court"), 1, 0),   # Either "Federal court" OR "State court"
         mid_term = ifelse(Year %% 4 == 2, 1, 0))

redistricting %>%
  select(Year, State, Drawn.by, legislature, partisan, independent_commission, court, mid_term)

```

\

## Clean Election data 


Keep only for required span - 2000 to 2020
Drop states with only one Congressional district. 
Then, convert the party variable to have only three categories - Republican, Democrat and Other. Pivot to wide, and get vote totals and shares.

Groupby/ aggregate and summarize district-level data to state level data.

\

```{r , echo=TRUE, message =FALSE, warning=FALSE}
# Keep only for required span - 2000 to 2020
  # drop  single dist states  
    
  # Vary from year?
single_dists <- c("AK", "DE", "MT", "ND", "SD", "VT", "WY")

elec <- elec %>%
  filter(year >= 2000) %>%
  filter(!state_po %in% single_dists)


############################################################
# Aggregation

# Generate new party2 variable

elec <- elec %>%
  mutate(party2 = case_when(
    party =="DEMOCRAT" ~ 'DEMOCRAT',
    party=="REPUBLICAN" ~ "REPUBLICAN",
    TRUE ~ as.character('OTHER'))) 


# Pivot wide - column for party totals
votes <- elec %>% pivot_wider(names_from = party2, values_from = candidatevotes) %>%
  mutate(across(c(DEMOCRAT, REPUBLICAN, OTHER), ~case_when(.== "NULL" ~ 0, 
                                  TRUE ~as.numeric(as.character(.)))))

# total turnout column
votes$turnout <- votes$DEMOCRAT + votes$REPUBLICAN + votes$OTHER

# Data now is at State-Year-Many districts - 
# aggregate statistics for all districts to one observation per state

votes <- votes %>%  
  group_by(state_po, year) %>% 
  summarize(demv = sum(DEMOCRAT), repv = sum(REPUBLICAN), othv = sum(OTHER), turnout = sum(turnout)) 

# create new "map_year" column in elec df as year-1 rounded to 10s
# the year in which the map for that election was drawn
votes <- votes %>%
  mutate(map_year = case_when(
              year <= 2010 ~ 2000,
              year > 2010 ~ 2010))

```

# Merge 


Merge the *Redistricting* and *Votes* datasets, on two keys: State and Map-Year

```{r , echo=TRUE, message =FALSE, warning=FALSE}

df <- votes %>% 
  left_join(redistricting, by = c("state_po" = "State", "map_year" = "Cycle.Year"))

# Cleaning for regression

df <- df %>%
  mutate(othv2=ifelse(is.na(othv), 0, othv))

df$turnout2 <- df$othv2 + df$demv + df$repv

```


# Download census data

https://rconsortium.github.io/censusguide/r-packages-all.html

https://walker-data.com/tidycensus/articles/spatial-data.html


The following code uses the tidycensus package and the Census API to download data + shapefiles from the US Census.

- Declare Census product (ACS, decennial), variables, time and geography.

- Reshape to wide, rename variables.

- Subset to US mainland; filter out Alaska, Hawaii, and Puerto Rico.

- Generate a variable of percent black

- Plot map


```{r , echo=TRUE, message=FALSE, warning=FALSE}
# Declare parameters - API key, variables, time
census_api_key(keyring::key_get("census_api_key"))

race_vars <- c("P003001", "P003002", "P003003")

yr_select <- c(2010)

allst <- get_decennial(geography = "state", 
                        variables = race_vars, 
                         geometry = TRUE,
                        year = yr_select) 

#idvars = c("NAME", "GEOID", "geometry")

# SPREAD to wide
allst2 <- spread(allst, variable, value)

setnames(allst2, old = c("P003001", "P003002", "P003003"), new = c('tot_pop','white_pop','black_pop'))

allst2 <- allst2 %>% 
  filter(NAME!="Puerto Rico", NAME!="Alaska", NAME!="Hawaii")

allst2$black_perc <- allst2$black_pop/allst2$tot_pop


allst2 %>% ggplot(aes(fill = black_perc)) + 
  geom_sf(color = NA) + 
  scale_fill_viridis_c(option = "viridis") 
```

## Join Eligible voters data 
Census data was available for the current census and historical data (separated).
Current data is combined, historical data is by presidential/midterms.
We import the relevant cells from the Excel sheets and merge them as needed.

```{r}

# 2020 file has raw populations (others have only percentages)
raw_2020 <- read_excel("../source_data/VoteAndRegByState2020.xlsx", range = "A7:E58")
names(raw_2020) <- c("State", "population", "citizens", "registered", "registered_percent")

# Presidential years (in 'broken' format in excel sheet (printer friendly))
rng <- "A7:K58"
prez_wide_regA <- read_excel("../source_data/VoteAndRegByStateHistory_prez_reg.xlsx", range=rng)
rng <- "B74:C125"
prez_wide_regB <- read_excel("../source_data/VoteAndRegByStateHistory_prez_reg.xlsx", range=rng)

prez_wide <- data.frame(cbind(prez_wide_regA, prez_wide_regB)[,seq(1,14,2)])
names(prez_wide) <- c("turnout_state", c(2020, 2016, 2012, 2008, 2004, 2000))

# MidTerms years (not in 'broken' format)
rng <- "A7:K58"
midt_wide_reg <- read_excel("../source_data/VoteAndRegByStateHistory_congress.xlsx", range=rng)
# rng <- "B74:C124"
# prez_wide_regB <- read_excel("../source_data/VoteAndRegByStateHistory_prez_reg.xlsx", range=rng)

midt_wide <- data.frame(midt_wide_reg[,seq(1,12,2)])
names(midt_wide) <- c("turnout_state", c(2018, 2014, 2010, 2006, 2002))

all_years <- left_join(prez_wide, midt_wide,
                       by = c("turnout_state" = "turnout_state"))

all_years_long <- pivot_longer(all_years, cols=colnames(all_years)[-1], names_to="turnout_year", values_to="turnout_perc")


```



# Descriptive results

Redistricting MAP

Vote shares and competitiveness - MAP/LINEPLOT

Census characteristics - AVERAGES?


```{r , echo=TRUE, message =FALSE, warning=FALSE}


```

# Regressions

Turnout percent into 

``` {r}


```





